---
title: "6303: Final Project"
author: "Varun Sai Thota 1002140673, Ramyasai Donkeshwaram 1002164699"
date: "2024-05-08"
output:
  pdf_document: default
  html_document: default
  word_document: default
---
# Loading required libraries
```{r}
library(dplyr)
library(caret)
library(ggplot2)
library(reshape2)
library(rpart.plot)
library(pROC)
```

# Load and view the recruitment dataset.
```{r}
recruitmentdataset <- read.csv("recruitmentdataset.csv")
View(recruitmentdataset)
```

# Display the first 5 rows of the recruitment dataset.
```{r}
head(recruitmentdataset, n = 5)
```
# Display the last 5 rows of the recruitment dataset.
```{r}
tail(recruitmentdataset, n = 5)
```

# Remove the 'Id' column from the dataset and display the first 5 rows of the updated dataset.
```{r}
recruitmentdataset$Id <- NULL
head(recruitmentdataset, n = 5)
```

# Display the dimensions of the dataset, remove duplicate rows, and display the new dimensions of the updated dataset.
```{r}
dim(recruitmentdataset)
recruitmentdataset <- unique(recruitmentdataset)
dim(recruitmentdataset)
```

```{r}
# Total Null values
total_nulls <- sum(is.na(recruitmentdataset))
print(total_nulls)
```

```{r}
# Summary of data
summary(recruitmentdataset)
```

# Checking data types 
```{r}
# Use str() function to display the structure of the dataset
str(recruitmentdataset)
```

```{r}
sapply(recruitmentdataset, class)
```
# Extract the names of all numeric columns in the recruitment dataset.
```{r}
numeric_cols <- recruitmentdataset[, sapply(recruitmentdataset, is.numeric)]
numeric_cols_names <- colnames(numeric_cols)
numeric_cols_names
```

# Outliers 

```{r}
# Visualizing class differences for numerical features

for (feature in numeric_cols_names) {
  p <- ggplot(recruitmentdataset, aes(y = !!sym(feature))) +
    geom_boxplot(fill = "steelblue", color = "black") +
    labs(title = paste("Box plot for", feature)) +
    theme_minimal() 
  print(p)
}
```

# Display the first rows of numeric columns using dplyr.
```{r}
library(dplyr)
head(select_if(recruitmentdataset, is.numeric))

```
# Get and display the names of numeric columns in the dataset.
```{r}
numeric_cols <- recruitmentdataset[, sapply(recruitmentdataset, is.numeric)]
numeric_cols_names <- colnames(numeric_cols)
numeric_cols_names
```

# Plot 1 ------ box plot for each numerical column against decision

```{r}
# Visualizing class differences for numerical features

for (feature in numeric_cols_names) {
  p <- ggplot(recruitmentdataset, aes(x = decision, y = !!sym(feature))) +
    geom_boxplot(fill = "steelblue", color = "black") +
    labs(title = paste("decision Vs", feature),
         x = "decision",
         y = feature) +
    theme_minimal() 
  print(p)
}
```

# Plot 2 --- Box plot for numerical columns
```{r}
# Visualizing class differences for numerical features

for (feature in numeric_cols_names) {
  p <- ggplot(recruitmentdataset, aes(y = !!sym(feature))) +
    geom_boxplot(fill = "steelblue", color = "black") +
    labs(title = paste("Box plot for", feature)) +
    theme_minimal() 
  print(p)
}
```

# Plot 3 --- Correlation Plot 

```{r}
library(reshape2)
# Select numerical variables for the correlation matrix
numeric_cols <- recruitmentdataset[, sapply(recruitmentdataset, is.numeric)]
numeric_cols_names <- colnames(numeric_cols)
numeric_cols_names

# Creating the correlation matrix
corr_matrix <- cor(numeric_cols)
# Converting the correlation matrix into a data frame
corr_dataframe <- melt(corr_matrix)

# Creating heatmap
ggplot(corr_dataframe, aes(Var2, Var1, fill = abs(value), label = round(value, 2))) +
  geom_tile(color = "white") +
  geom_text(size = 3, color = "black") +  # Ajouter les valeurs de corrélation
  scale_fill_gradientn(colors = c("white","steelblue", "blue"), name="Correlation") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 10, hjust = 1)) +
  coord_fixed() +
  labs(x = "", y = "") +
  theme(legend.position = "right")
```

```{r}
head(recruitmentdataset)
```

# Plot 4 ---- Violin plot for income and age

```{r}
library(ggplot2)
# Most basic violin chart
p <- ggplot(recruitmentdataset, aes(x=decision, y=age, fill=decision)) + # fill=name allow to automatically dedicate a color for each group
  geom_violin()
p
```

# Plot 5 ---- Bar Chart - Programming Experience by Recruitment Decision

```{r}
library(ggplot2)
# Convert 'decision' to a factor
recruitmentdataset$decision <- as.factor(recruitmentdataset$decision)
# Ensure levels of decision match True/False
levels(recruitmentdataset$decision) <- c("False", "True")
# Bar chart for Programming Experience by Decision
ggplot(recruitmentdataset, aes(x = ind.programming_exp, fill = decision)) +
  geom_bar(position = "dodge") +
  ggtitle("Programming Experience by Recruitment Decision") +
  xlab("Programming Experience (Years)") +
  ylab("Count of Candidates") +
  theme_minimal() +
  scale_fill_manual(values = c("False" = "red", "True" = "blue"), 
                    name = "Decision",
                    labels = c("Not Hired", "Hired"))


```

# Plot 6 ---- Bar Chart - Degree Levels by Recruitment Decision
```{r}
library(ggplot2)
# Convert 'decision' to a factor
recruitmentdataset$decision <- as.factor(recruitmentdataset$decision)
# Ensure levels of decision match True/False
levels(recruitmentdataset$decision) <- c("False", "True")
# Degree vs Decision
ggplot(recruitmentdataset, aes(x = ind.degree , fill = decision)) +
  geom_bar(position = "dodge") +
  ggtitle("Degree Levels by Recruitment Decision") +
  xlab("Degree Type") +
  ylab("Count of Candidates") +
  theme_minimal() +
  scale_fill_manual(values = c("False" = "red", "True" = "blue"), 
                    name = "Decision",
                    labels = c("Not Hired", "Hired"))
```

# Decision trees(Gini index and Gain ratio)

```{r}
# Load necessary libraries
library(caret)
library(rpart)
library(rpart.plot)
library(ROCR)
library(ggplot2)
# Data Partitioning for Gini Index
set.seed(4699)
trainIndex <- createDataPartition(recruitmentdataset$decision, p = 0.80, list = FALSE)
train.data1 <- recruitmentdataset[trainIndex, ]
test.data1 <- recruitmentdataset[-trainIndex, ]
cat("Training Data Dimensions (Gini):", dim(train.data1), "\n")
cat("Testing Data Dimensions (Gini):", dim(test.data1), "\n")

# Decision Tree using Gini Index
tree_gini <- rpart(decision ~ ., data = train.data1, method = "class", parms = list(split = "gini"))

# Plot Decision Tree with Gini Index
rpart.plot(tree_gini, main = "Decision Tree (Gini Index)", type = 4, extra = 101)

# Data Partitioning for Gain Ratio
set.seed(673)
trainIndex <- createDataPartition(recruitmentdataset$decision, p = 0.80, list = FALSE)
train.data2 <- recruitmentdataset[trainIndex, ]
test.data2 <- recruitmentdataset[-trainIndex, ]
cat("Training Data Dimensions (Gain Ratio):", dim(train.data2), "\n")
cat("Testing Data Dimensions (Gain Ratio):", dim(test.data2), "\n")

# Decision Tree using Gain Ratio
tree_gain_ratio <- rpart(decision ~ ., data = train.data2, method = "class", parms = list(split = "information"))

# Plot Decision Tree with Gain Ratio
rpart.plot(tree_gain_ratio, main = "Decision Tree (Gain Ratio)", type = 4, extra = 101)

# Cross-Validation Control
trctrl <- trainControl(method = "cv", classProbs = TRUE)

# Fit Decision Tree Classification Model (Gini Index)
DTModel_gini <- train(decision ~ ., data = train.data1, method = "rpart", metric = "ROC", parms = list(split = "gini"), trControl = trctrl)
cat("Model Summary (Gini Index):\n")
print(DTModel_gini)

# Fit Decision Tree Classification Model (Gain Ratio)
DTModel_gain_ratio <- train(decision ~ ., data = train.data2, method = "rpart", metric = "ROC", parms = list(split = "information"), trControl = trctrl)
cat("Model Summary (Gain Ratio):\n")
print(DTModel_gain_ratio)

# Visualization for Gini Index
prp(DTModel_gini$finalModel, box.palette = "orange", tweak = 1.2, varlen = 20)

# Visualization for Gain Ratio
prp(DTModel_gain_ratio$finalModel, box.palette = "yellow", tweak = 1.2, varlen = 20)

# Variable Importance for Gini Index
model_importance_gini <- varImp(DTModel_gini)
importance_df_gini <- as.data.frame(model_importance_gini$importance)
importance_df_gini <- cbind(variable_name = rownames(importance_df_gini), importance_df_gini)
rownames(importance_df_gini) <- 1:nrow(importance_df_gini)
sorted_df_gini <- importance_df_gini[order(-importance_df_gini$Overall), ]
top_variables_gini <- head(sorted_df_gini, 8)

# Plot Variable Importance (Gini)
ggplot(top_variables_gini, aes(x = Overall, y = reorder(variable_name, Overall), fill = Overall)) +
  geom_bar(stat = "identity") +
  scale_fill_gradient2(low = "skyblue", high = "yellow", midpoint = 50) +
  labs(x = "Overall Importance Score", y = "Variable", title = "Variable Importance - Gini") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# Variable Importance for Gain Ratio
model_importance_gain_ratio <- varImp(DTModel_gain_ratio)
importance_df_gain_ratio <- as.data.frame(model_importance_gain_ratio$importance)
importance_df_gain_ratio <- cbind(variable_name = rownames(importance_df_gain_ratio), importance_df_gain_ratio)
rownames(importance_df_gain_ratio) <- 1:nrow(importance_df_gain_ratio)
sorted_df_gain_ratio <- importance_df_gain_ratio[order(-importance_df_gain_ratio$Overall), ]
top_variables_gain_ratio <- head(sorted_df_gain_ratio, 8)

# Plot Variable Importance (Gain Ratio)
ggplot(top_variables_gain_ratio, aes(x = Overall, y = reorder(variable_name, Overall), fill = Overall)) +
  geom_bar(stat = "identity") +
  scale_fill_gradient2(low = "lightpink", high = "red", midpoint = 50) +
  labs(x = "Overall Importance Score", y = "Variable", title = "Variable Importance - Gain Ratio") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# Confusion Matrix for Gini Index
gini_predictions <- predict(DTModel_gini, test.data1, type = "raw")
gini_conf_matrix <- confusionMatrix(gini_predictions, test.data1$decision)
cat("Confusion Matrix for Gini Index:\n")
print(gini_conf_matrix)

# Confusion Matrix for Gain Ratio
gain_ratio_predictions <- predict(DTModel_gain_ratio, test.data2, type = "raw")
gain_ratio_conf_matrix <- confusionMatrix(gain_ratio_predictions, test.data2$decision)
cat("Confusion Matrix for Gain Ratio:\n")
print(gain_ratio_conf_matrix)

# AUC-ROC for Gini Index
gini_probabilities <- predict(DTModel_gini, test.data1, type = "prob")[, 2]
roc_curve_gini <- roc(test.data1$decision, gini_probabilities)
auc_value_gini <- auc(roc_curve_gini)
cat("AUC for Gini Index:", auc_value_gini, "\n")

plot(roc_curve_gini, main = "AUC-ROC Curve (Gini Index)", col = "blue")

# AUC-ROC for Gain Ratio
gain_ratio_probabilities <- predict(DTModel_gain_ratio, test.data2, type = "prob")[, 2]
roc_curve_gain_ratio <- roc(test.data2$decision, gain_ratio_probabilities)
auc_value_gain_ratio <- auc(roc_curve_gain_ratio)
cat("AUC for Gain Ratio:", auc_value_gain_ratio, "\n")

plot(roc_curve_gain_ratio, main = "AUC-ROC Curve (Gain Ratio)", col = "red")

```

# Random Forest

```{r}
# Load necessary libraries
library(randomForest)
library(caret)
library(pROC)
library(ggplot2)

# Set seed for reproducibility
set.seed(123)

# Split the data into training (80%) and testing (20%) subsets
split_index <- createDataPartition(recruitmentdataset$decision, p = 0.8, list = FALSE)
train_data <- recruitmentdataset[split_index, ]
test_data <- recruitmentdataset[-split_index, ]

# Ensure the target variable is a factor with valid R variable names for levels
train_data$decision <- factor(train_data$decision, levels = unique(train_data$decision))
levels(train_data$decision) <- make.names(levels(train_data$decision))

test_data$decision <- factor(test_data$decision, levels = unique(test_data$decision))
levels(test_data$decision) <- make.names(levels(test_data$decision))

# Perform Cross-Validation to Tune Hyperparameters
set.seed(123)
cv_control <- trainControl(method = "cv", number = 10, classProbs = TRUE, summaryFunction = twoClassSummary)

# Tune mtry using cross-validation
rf_cv <- train(
  decision ~ ., 
  data = train_data, 
  method = "rf", 
  metric = "ROC", 
  tuneGrid = expand.grid(mtry = c(2, 5, 10, 13)), # Test different mtry values
  trControl = cv_control,
  ntree = 500
)

# Print Cross-Validation Results
print(rf_cv)

# Extract the best mtry value
best_mtry <- rf_cv$bestTune$mtry
cat("Best mtry value from CV:", best_mtry, "\n")

# Train the Final Random Forest Model with Best mtry
random_forest_model <- randomForest(
  decision ~ ., 
  data = train_data, 
  ntree = 500, 
  mtry = best_mtry, 
  importance = TRUE
)

# Summary of the final model
print(random_forest_model)

# Evaluate variable importance
varImpPlot(random_forest_model, main = "Variable Importance Plot")

# Predict on the test dataset
rf_predictions <- predict(random_forest_model, newdata = test_data, type = "response")

# Confusion matrix to evaluate model performance
rf_conf_matrix <- confusionMatrix(rf_predictions, test_data$decision)
print(rf_conf_matrix)

# Add Kappa Score
rf_kappa_score <- rf_conf_matrix$overall['Kappa']
cat("Kappa Score: ", rf_kappa_score, "\n")

# Calculate and plot AUC-ROC
rf_probabilities <- predict(random_forest_model, newdata = test_data, type = "prob")[, 2]
rf_roc_curve <- roc(test_data$decision, rf_probabilities)
rf_auc_value <- auc(rf_roc_curve)
cat("Random Forest AUC: ", rf_auc_value, "\n")

plot(rf_roc_curve, main = "AUC-ROC Curve for Random Forest", col = "darkgreen")

# Save results for comparison
rf_results <- list(
  Model = "Random Forest",
  Accuracy = rf_conf_matrix$overall['Accuracy'],
  Kappa = rf_kappa_score,
  AUC = rf_auc_value
)
print(rf_results)

# Feature Importance Analysis
# Extract feature importance measures
rf_importance <- importance(random_forest_model)
rf_importance_df <- data.frame(
  Feature = rownames(rf_importance),
  MeanDecreaseAccuracy = rf_importance[, "MeanDecreaseAccuracy"],
  MeanDecreaseGini = rf_importance[, "MeanDecreaseGini"]
)

# Sort by Mean Decrease in Accuracy
rf_importance_df <- rf_importance_df[order(-rf_importance_df$MeanDecreaseAccuracy), ]

# Display top features
cat("Top Features for Random Forest (Mean Decrease Accuracy):\n")
print(head(rf_importance_df, 10))

# Plot feature importance (Mean Decrease Accuracy - Top 10 Features)
ggplot(rf_importance_df[1:10, ], aes(x = reorder(Feature, MeanDecreaseAccuracy), y = MeanDecreaseAccuracy)) +
  geom_bar(stat = "identity", fill = "forestgreen") +
  coord_flip() +
  ggtitle("Top 10 Feature Importance - Random Forests") +
  xlab("Features") +
  ylab("Mean Decrease Accuracy")

```

# XGboost

```{r}
# Load necessary libraries
library(xgboost)
library(caret)
library(pROC)

# Set seed for reproducibility
set.seed(123)

# Split the data into training (80%) and testing (20%) subsets
split_index <- createDataPartition(recruitmentdataset$decision, p = 0.8, list = FALSE)
train_data <- recruitmentdataset[split_index, ]
test_data <- recruitmentdataset[-split_index, ]

# Convert the target variable to binary (0/1)
train_data$decision <- as.numeric(as.factor(train_data$decision)) - 1
test_data$decision <- as.numeric(as.factor(test_data$decision)) - 1

# One-hot encode categorical variables
train_matrix <- model.matrix(decision ~ . - 1, data = train_data)
test_matrix <- model.matrix(decision ~ . - 1, data = test_data)

# Prepare DMatrix for XGBoost
dtrain <- xgb.DMatrix(data = train_matrix, label = train_data$decision)
dtest <- xgb.DMatrix(data = test_matrix, label = test_data$decision)

# Perform Cross-Validation
cv_results <- xgb.cv(
  data = dtrain,
  nrounds = 100,               # Number of boosting rounds
  nfold = 10,                  # 10-fold cross-validation
  objective = "binary:logistic",
  eval_metric = "auc",
  verbose = 0,                 # Suppress training output
  early_stopping_rounds = 10,  # Stop if no improvement after 10 rounds
  print_every_n = 10           # Print every 10 iterations
)

# Extract and print the best iteration
best_iteration <- cv_results$best_iteration
cat("Best number of boosting rounds (best iteration):", best_iteration, "\n")

# Train an XGBoost model with default parameters
xgb_model <- xgboost(
  data = dtrain,
  objective = "binary:logistic",
  eval_metric = "auc",
  nrounds = cv_results$best_iteration, # Use the best number of iterations from CV
  verbose = 0                          # Suppress training output
)

# Predict on the test dataset
xgb_probabilities <- predict(xgb_model, newdata = dtest)

# Convert probabilities to class labels (using a threshold of 0.5)
xgb_predictions <- ifelse(xgb_probabilities > 0.5, 1, 0)

# Confusion matrix to evaluate model performance
xgb_conf_matrix <- confusionMatrix(as.factor(xgb_predictions), as.factor(test_data$decision))
print(xgb_conf_matrix)

# Calculate and plot AUC-ROC
xgb_roc_curve <- roc(test_data$decision, xgb_probabilities)
xgb_auc_value <- auc(xgb_roc_curve)
cat("XGBoost AUC: ", xgb_auc_value, "\n")

plot(xgb_roc_curve, main = "AUC-ROC Curve for XGBoost", col = "red")

# Feature Importance
importance <- xgb.importance(model = xgb_model)
cat("Feature Importance:\n")
print(importance)

# Plot Feature Importance (Top 10 Features)
xgb.plot.importance(importance, top_n = 10, main = "Top 10 Feature Importance")

```

